Sign language serves as the primary mode of communication for over 
70 million deaf and hard-of-hearing individuals worldwide \cite{who2021}. 
American Sign Language (ASL), used predominantly in North America, employs a 
rich vocabulary of hand gestures and finger-spelling to convey words and concepts. 
Despite its widespread use, the absence of real-time automated translation tools 
remains a critical barrier in everyday communication with the hearing majority.

Recent advances in convolutional neural networks (CNNs) and real-time object 
detection frameworks have opened new avenues for gesture recognition. Systems 
that can process live video frames and return a classified gesture label within 
milliseconds represent a meaningful step toward accessible communication technology.
In the first stage, a MediaPipe hand tracking module localizes the hand region within each 
camera frame, producing a tight bounding box crop(invisible to the user) and a skeleton tracking system(visible). 
In the second stage, a fine-tuned 
ResNet-50 \cite{he2016} backbone classifies the cropped hand image into one of 26 ASL 
alphabet letters. The result is overlaid on the live video stream in real time.

Our contributions are as follows:
\begin{itemize}
    \item A modular training pipeline with configurable optimizer, scheduler, and early stopping, implemented in PyTorch.
    \item Integration of MediaPipe-based hand detection with a ResNet-50 classification head for end-to-end ASL letter recognition.
    \item A real-time inference module with webcam integration and visual feedback.
    \item A FastAPI web interface enabling browser-based interaction with the live pipeline.
\end{itemize}