\documentclass[conference]{IEEEtran}
\IEEEoverridecommandlockouts

\usepackage{cite}
\usepackage{amsmath,amssymb,amsfonts}
\usepackage{algorithmic}
\usepackage{graphicx}
\usepackage{textcomp}
\usepackage{xcolor}
\usepackage{hyperref}
\usepackage{booktabs}
\usepackage{array}

\def\BibTeX{{\rm B\kern-.05em{\sc i\kern-.025em b}\kern-.08em
    T\kern-.1667em\lower.7ex\hbox{E}\kern-.125emX}}

\begin{document}

\title{Real-Time American Sign Language Detection Using MediaPipe Hand 
Localization and ResNet-50 Classification}

\author{
  \IEEEauthorblockN{Udit Asthana}
  \IEEEauthorblockA{\textit{Manipal Institute of Technology} \\ Reg. No: 240905310 \\ Roll No: 31}
  \and
  \IEEEauthorblockN{Abhishek L Prabhu}
  \IEEEauthorblockA{\textit{Manipal Institute of Technology} \\ Reg. No: 240905676 \\ Roll No: 71}
  \and
  \IEEEauthorblockN{Ashwath Patil}
  \IEEEauthorblockA{\textit{Manipal Institute of Technology} \\ Reg. No: 240905460 \\ Roll No: 46}
  \and 
  \IEEEauthorblockN{Kishan Kanvathirtha}
  \IEEEauthorblockA{\textit{Manipal Institute of Technology} \\ Reg. No: 240905678 \\ Roll No: 72}
  \and
  \IEEEauthorblockN{Arnav Dugad}
  \IEEEauthorblockA{\textit{Manipal Institute of Technology} \\ Reg. No: 230905227 \\ Roll No: 73}
}

\maketitle

\begin{abstract}
American Sign Language (ASL) is the primary mode of non-verbal communication for 
the hearing-impaired community. Automated real-time recognition of ASL gestures 
can bridge the communication gap between hearing-impaired and hearing individuals. 
This paper proposes a two-stage pipeline for real-time ASL alphabet recognition: 
hand region localization using MediaPipe, followed by letter classification using a 
fine-tuned ResNet-50 backbone with a custom classification head. 
The system is trained on the publicly available ASL Alphabet dataset 
from Kaggle and evaluated across 26 letter classes. A cosine annealing 
learning rate scheduler with linear warm-up, gradient clipping, and early stopping are 
employed to stabilize training. The proposed system achieves competitive accuracy and 
operates in real-time via a live webcam feed, with bounding box overlays and predicted 
letter labels rendered on each frame. A lightweight FastAPI-based frontend serves the 
inference pipeline for interactive use. This work demonstrates the viability of 
combining detection and classification CNNs for low-latency sign language recognition.
\end{abstract}

\begin{IEEEkeywords}
sign language recognition, ASL, MediaPipe, ResNet-50, hand detection, 
real-time inference, deep learning, convolutional neural networks
\end{IEEEkeywords}

% -------------------------------------------------------
\section{Introduction}

Sign language serves as the primary mode of communication for over 
70 million deaf and hard-of-hearing individuals worldwide \cite{who2021}. 
American Sign Language (ASL), used predominantly in North America, employs a 
rich vocabulary of hand gestures and finger-spelling to convey words and concepts. 
Despite its widespread use, the absence of real-time automated translation tools 
remains a critical barrier in everyday communication with the hearing majority.

Recent advances in convolutional neural networks (CNNs) and real-time object 
detection frameworks have opened new avenues for gesture recognition. Systems 
that can process live video frames and return a classified gesture label within 
milliseconds represent a meaningful step toward accessible communication technology.
aa MediaPipe object detector localizes the hand region within each 
camera fa MediaPipe hand tracking module localizes the hand region within each 
camera frame, producing a tight bounding box crop. In the second stage, a fine-tuned 
ResNet-50 \cite{he2016} backbone classifies the cropped hand image into one of 26 ASL 
alphabet letters. The result is overlaid on the live video stream in real time.

Our contributions are as follows:
\begin{itemize}
    \item A modular training pipeline with configurable optimizer, scheduler, and early stopping, implemented in PyTorch.
    \item Integration of MediaPipe-based hand detection with a ResNet-50 classification head for end-to-end ASL letter recognition.
    \item A real-time inference module with webcam integration and visual feedback.
    \item A FastAPI web interface enabling browser-based interaction with the live pipeline.
\end{itemize}

% -------------------------------------------------------
\section{Related Work}

Early sign language recognition systems relied heavily on sensor-based approaches, such as data gloves and depth cameras, to capture 3D hand pose information \cite{mitra2007}. While effective, these methods impose hardware constraints that limit practical deployment.

With the rise of deep learning, image-based approaches became dominant. Convolutional networks trained on static hand images have demonstrated strong classification performance on isolated gestures \cite{pigou2015}. Transfer learning using ImageNet-pretrained backbones such as VGG-16, ResNet, and MobileNet has further improved accuracy on smaller, domain-specific datasets.

Real-time landmark-based real-time hand tracking systems have been explored for gesture spotting, but comprehensive two-stage systems targeting the full ASL alphabet in a live-camera setting remain relatively underexplored. This work builds on these foundations to deliver a practical, deployable system.

% -------------------------------------------------------
\section{Dataset}

The ASL Alphabet dataset from Kaggle \cite{kaggle_asl} contains 87,000 labeled images spanning 29 classes: 26 letters (A–Z), plus \texttt{SPACE}, \texttt{DELETE}, and \texttt{NOTHING}. Each image is $200 \times 200$ pixels in RGB format.

For this work, the dataset is restricted to the 26 alphabetic gesture classes. Images are resized to $50 \times 50$ pixels and converted to tensors for input to the ResNet-50 backbone. The dataset is split into training (70\%), validation (20\%), and test (10\%) subsets using stratified random splitting to preserve class balance. Batch size is set to 512 with shuffling enabled for training and disabled for evaluation.

% -------------------------------------------------------
\section{Methodology}

\subsection{System Overview}

The proposed pipeline operates in two sequential stages at inference time:

\begin{enumerate}
    \item \textbf{Hand Detection (MediaPipe):} Each incoming video frame is passed to a MediaPipe model that produces bounding boxes around detected hand regions.
    \item \textbf{Letter Classification (ResNet-50):} The detected hand crop is extracted, resized, and passed to the classification model, which returns a probability distribution over 26 ASL letter classes.
\end{enumerate}

The predicted letter and bounding box are overlaid on the original frame for visual feedback.

\subsection{Hand Detection Module}

MediaPipe Hands is used for real-time hand localization and tracking. 
Unlike traditional object detection frameworks, MediaPipe employs a 
two-stage pipeline consisting of a palm detector followed by a hand 
landmark model. The system directly predicts 21 2D keypoints corresponding 
to anatomical hand joints, along with a bounding region derived from these 
landmarks.

Given an input frame, MediaPipe returns:
\begin{itemize}
    \item 21 hand landmarks (normalized coordinates)
    \item Handedness (left/right classification)
    \item A region of interest (ROI) derived from landmark extents
\end{itemize}

This approach provides fine-grained spatial information about finger 
articulation, enabling robust downstream classification even in cluttered 
backgrounds. The ROI extracted from the landmarks is used as input to the 
classification network.

.

\subsection{Classification Model (SLD)}
ets.

\subsection{Classification Model (SLD)}

The Sign Language Detector (\textbf{SLD}) model consists of a frozen ResNet-50 backbone followed by a lightweight classification head:

\begin{equation}
    \hat{y} = \text{head}(\text{backbone}(x))
\end{equation}

where the backbone extracts a 2048-dimensional feature vector and the head is defined as:

\begin{equation}
    \text{head}(z) = W_2 \cdot \text{GELU}(\text{Dropout}(\text{LayerNorm}(z) \cdot W_1))
\end{equation}

with $W_1 \in \mathbb{R}^{2048 \times 512}$ and $W_2 \in \mathbb{R}^{512 \times 26}$.

\textbf{LayerNorm} replaces BatchNorm at the boundary between backbone and head to stabilize fine-tuning. \textbf{GELU} activation is chosen for its smoothness properties. \textbf{Dropout} ($p = 0.3$) regularizes the intermediate representation.

\subsection{Training Configuration}

Training is governed by a dataclass-based configuration (\texttt{TRAINING\_CONFIG}) that encapsulates all hyperparameters for reproducibility. Key settings are summarized in Table~\ref{tab:config}.

\begin{table}[h]
\caption{Training Hyperparameters}
\label{tab:config}
\centering
\begin{tabular}{ll}
\toprule
\textbf{Parameter} & \textbf{Value} \\
\midrule
Optimizer & SGD (Nesterov) \\
Learning Rate & $1 \times 10^{-2}$ \\
Momentum & 0.9 \\
Weight Decay & $5 \times 10^{-4}$ \\
Scheduler & Cosine Annealing + Warm-Up \\
Warm-Up Epochs & 10 \\
$T_{\max}$ & 200 \\
$\eta_{\min}$ & $1 \times 10^{-4}$ \\
Max Epochs & 200 \\
Batch Size & 512 \\
Gradient Clip Norm & 1.0 \\
Early Stopping Patience & 20 \\
\bottomrule
\end{tabular}
\end{table}

\subsubsection{Learning Rate Schedule}

A cosine annealing schedule with linear warm-up is used. During the warm-up phase (epochs $0$ to $T_w - 1$), the learning rate increases linearly:

\begin{equation}
    \eta_t = \eta_{\text{base}} \cdot \frac{t+1}{T_w}
\end{equation}

After warm-up, cosine decay is applied:

\begin{equation}
    \eta_t = \eta_{\min} + \frac{\eta_{\text{base}} - \eta_{\min}}{2} \left(1 + \cos\left(\pi \cdot \frac{t - T_w}{T_{\max} - T_w}\right)\right)
\end{equation}

This prevents early training instability while allowing the learning rate to settle to a low value near convergence.

\subsubsection{Gradient Clipping}

Gradient norm clipping at threshold $\tau = 1.0$ is applied after each backward pass to prevent exploding gradients:

\begin{equation}
    \mathbf{g} \leftarrow \mathbf{g} \cdot \min\!\left(1,\ \frac{\tau}{\|\mathbf{g}\|_2}\right)
\end{equation}

\subsubsection{Early Stopping}

Training is halted if validation accuracy does not improve for 20 consecutive epochs. The best-performing checkpoint is saved and loaded for final test evaluation.

\subsection{Experiment Tracking}

All training runs are logged via Weights \& Biases (W\&B) \cite{wandb}, tracking per-epoch training loss, validation loss, accuracy, learning rate, mean gradient norm, and clipping ratio.

% -------------------------------------------------------
\section{Real-Time Inference Pipeline}

At inference time, frames are captured via OpenCV from a live webcam feed. Each frame undergoes the following steps:

\begin{enMediaPipe
    \item \textbf{Hand Detection:} MediaPipe processes the BGR frame and returns bounding boxes.
    \item \textbf{Crop and Resize:} The highest-confidence bounding box crop is resized to $50 \times 50$ and normalized.
    \item \textbf{Classification:} The crop is passed through the trained SLD model to obtain a predicted letter.
    \item \textbf{Overlay:} The bounding box and predicted label are drawn onto the original frame using OpenCV.
    \item \textbf{Display:} The annotated frame is rendered in real time.
\end{enumerate}

\subsection{Frontend API (Planned)}

A lightweight FastAPI \cite{fastapi} server will expose the inference pipeline as a REST API. The planned endpoints are:

\begin{itemize}
    \item \texttt{POST /predict} — accepts a base64-encoded image frame and returns the predicted letter and bounding box.
    \item \texttt{GET /stream} — serves a server-sent event stream of predictions from a webcam.
    \item \texttt{GET /health} — returns model and system status.
\end{itemize}

A minimal browser-based frontend will display the live annotated stream and maintain a rolling transcript of predicted letters, enabling real-time spell-out of words and sentences.

% -------------------------------------------------------
\section{Results and Discussion}

\textit{[Results to be updated with final trained model metrics in the Final Draft.]}

The model is currently under training. Preliminary evaluation metrics — test accuracy, per-class precision, recall, and F1-score — will be reported in the final submission. A confusion matrix across all 26 ASL classes will be included to analyze inter-class ambiguity (e.g., M/N, R/U which are visually similar in static images).

We anticipate competitive classification performance given the depth of the ResNet-50 backbone and the quality of the Kaggle ASL dataset. The primary challenge is generalization to real-world hand appearances under varying lighting and background conditions not represented in the training set.

% -------------------------------------------------------
\section{Conclusion}

This paper presents a real-tMediaPipe8phabet recognition system combining MediaPipe hand detection with ResNet-50 classification. The system is trained with a robust pipeline incorporating warm-up scheduling, gradient clipping, and early stopping, and is designed for live deployment via a webcam and FastAPI interface. This work addresses a meaningful accessibility problem and establishes a practical baseline for further extensions, including full ASL word-level recognition and continuous signing.

Future directions incMediaPipe8gin{itemize}
    \item Fine-tuning MediaPipe on a dedicated hand detection dataset for improved localization accuracy.
    \item Upgrading the backbone to ResNet-152 or EfficientNet for higher classification capacity.
    \item Extending the vocabulary to dynamic gestures and full ASL words using temporal models (e.g., LSTM or Transformer-based sequence models).
    \item Deploying the system as a mobile application for broader accessibility.
\end{itemize}

% -------------------------------------------------------
\begin{thebibliography}{00}

\bibitem{who2021}
World Health Organization, ``Deafness and hearing loss,'' \textit{WHO Fact Sheets}, 2021. [Online]. Available: \url{https://www.who.int/news-room/fact-sheets/detail/deafness-and-hearing-loss}

\bibitem{redmon2016}
J. Redmon, S. Divvala, R. Girshick, and A. Farhadi, ``You only look once: Unified, real-time object detection,'' in \textit{Proc. IEEE CVPR}, 2016, pp. 779--788.

\bibitem{he2016}
K. He, X. Zhang, S. Ren, and J. Sun, ``Deep residual learning for image recognition,'' in \textit{Proc. IEEE CVPR}, 2016, pp. 770--778.

\bibitem{mitra2007}
S. Mitra and T. Acharya, ``Gesture recognition: A survey,'' \textit{IEEE Trans. Syst., Man, Cybern. C, Appl. Rev.}, vol. 37, no. 3, pp. 311--324, 2007.

\bibitem{pigou2015}
L. Pigou, S. Dieleman, P.-J. Kindermans, and B. Schrauwen, ``Sign language recognition using convolutional neural networks,'' in \textit{Proc. ECCV Workshops}, 2015.

\bibitem{kaggle_asl}
A. Akash, ``ASL Alphabet,'' Kaggle, 2018. [Online]. Available: \url{https://www.kaggle.com/datasets/grassknoted/asl-alphabet}

\bibitem{wandb}
L. Biewald, ``Experiment tracking with weights and biases,'' 2020. [Online]. Available: \url{https://www.wandb.com}

\bibitem{fastapi}
S. Ram\'irez, ``FastAPI,'' 2018. [Online]. Available: \url{https://fastapi.tiangolo.com}

\end{thebibliography}

\end{document}